---
title: "Resampling surface data with explicit coverage"
author: "Brad Buchsbaum"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Resampling surface data with explicit coverage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", message=FALSE)
library(neurotransform)
```

A surface plan maps values on a moving mesh to vertices of a reference mesh.
Build the geometry once, then apply it to vectors or matrices whose rows are
moving vertices. This example uses an octahedron and two query directions.

```{r ordinary}
vertices <- rbind(c(1,0,0), c(-1,0,0), c(0,1,0), c(0,-1,0),
                  c(0,0,1), c(0,0,-1))
faces <- rbind(c(1,3,5), c(3,2,5), c(2,4,5), c(4,1,5),
               c(3,1,6), c(2,3,6), c(4,2,6), c(1,4,6))
moving <- surface_mesh(vertices, faces)
query <- rbind(c(1,1,1)/sqrt(3), c(1,1,0)/sqrt(2))
reference <- surface_mesh(query)
plan <- surface_resampling_plan(reference, moving)
z <- vertices[,3]
apply_surface_resampling(plan, z)
```

The first result is one third: the positive-octant face has equal weights on
its three corners. The second query lies on an edge whose endpoints both have
value zero. Reordering faces leaves the interpolation unchanged.

```{r check-ordinary, include=FALSE}
stopifnot(isTRUE(all.equal(apply_surface_resampling(plan,z),c(1/3,0),tolerance=1e-12)))
```

Spherical plans normalize both geometries to a common radius about the origin
and select the closest point on the triangular mesh, including edges and
vertices. They require a closed, connected spherical source with an unambiguous
radial mapping. `validate_surface_mesh()` reports defects without repairing
input. A reference consisting only of query directions may omit faces.

For open, nonspherical surfaces, set `spherical=FALSE`. Barycentric plans then
use interior orthogonal projections; `outside="error"`, `"missing"`, and
`"nearest"` choose how uncovered queries are handled. The nonspherical default
is nearest-vertex fallback, which is identified separately in coverage details.
Direct `surface_sampler()` calls expose `projection="interior"` or `"closest"`.

## Distinguish missing values from zero

By default, a nonfinite value with positive weight makes that output unavailable.
A zero-weight missing value has no effect. Omission explicitly renormalizes the
remaining finite weights and reports how much raw weight remained.

```{r missing}
with_missing <- z
with_missing[5] <- NA_real_
apply_surface_resampling(plan, with_missing)
omitted <- apply_surface_resampling(plan, with_missing,
                                    na_policy="omit", details=TRUE)
data.frame(value=omitted$values,
           retained_weight=omitted$finite_weight_mass,
           available=omitted$available)
```

Omission returns zero at the first query, based only on the two equatorial
vertices; its retained weight is two thirds. An all-missing neighborhood returns
`NA`, not zero. With matrix data these diagnostics are column specific. Use
`na_policy="error"` to reject contributing nonfinite values. Omission requires
row normalization (`normalize="element"`).

Source and target ROIs have different roles. Positive source-mask entries
include vertices before row normalization; a target mask excludes outputs
after correspondence and normalization have been computed.

```{r masks}
masked_plan <- surface_resampling_plan(
  reference, moving,
  source_mask=c(1,1,1,1,0,0), target_mask=c(TRUE,FALSE)
)
coverage <- apply_surface_resampling(masked_plan, z, details=TRUE)
data.frame(value=coverage$values, geometry=coverage$geometric_support,
           source_weight=coverage$source_weight_mass, status=coverage$status)
```

Both queries have geometric support. The first keeps two thirds of its original
source weight and produces a supported zero. The second is unavailable because
the target mask excludes it. A topological hole in a spherical mesh is a geometry
error, not an ROI.

## Resample integer labels by voting

Labels use integer keys and an optional table of names or colors. Aggregate
voting sums weights for each key; largest voting chooses a single source vertex.
Neither policy averages the numeric IDs.

```{r labels}
keys <- c(10L,20L,10L,20L,30L,40L)
label_table <- data.frame(key=c(0L,10L,20L,30L,40L),
                          name=c("unassigned","A","B","C","D"))
apply_surface_resampling(plan, keys, data_type="label",
                         label_table=label_table, unassigned=0L)
```

Aggregate ties choose the smallest label key. Largest-weight ties choose the
smallest source vertex index. These are deterministic library conventions;
Workbench can resolve ties differently. Set `unassigned` explicitly when another
application uses a particular missing-label key, and include it in the label
table. The default is `NA_integer_`.

## Separate adjoint application from reverse resampling

For a fixed forward matrix `W`, `apply_surface_adjoint()` applies its Euclidean
transpose, including the forward normalization and fixed masks. It does not
renormalize the transpose.

```{r adjoint}
y <- c(2,1)
c(forward_inner_product=sum(apply_surface_resampling(plan,z)*y),
  adjoint_inner_product=sum(z*apply_surface_adjoint(plan,y)))
```

An unsupported source column has algebraic adjoint value zero; `details=TRUE`
identifies those columns. Missing-data omission changes weights with the data
and is not this fixed adjoint. `normalize="sum"` conserves represented discrete
source mass before target exclusion; it does not perform anatomical area
correction.

Reverse geometric interpolation requires faces on the new source. Supply both
original geometries so their recorded identities can be checked.

```{r reverse}
angle <- .23
rotation <- rbind(c(cos(angle),-sin(angle),0),
                  c(sin(angle),cos(angle),0),c(0,0,1))
rotated_vertices <- vertices %*% rotation
rotated <- surface_mesh(rotated_vertices, faces)
forward_plan <- surface_resampling_plan(rotated, moving)
reverse_plan <- reverse_surface_resampling_plan(forward_plan, rotated, moving)
reverse_plan
```

A rebuilt reverse plan generally differs from a transpose and cannot recover
information lost through downsampling. `inverse=TRUE` is deprecated and retains
its legacy transpose-then-normalize behavior; it rejects new policies and fixed
masks rather than silently changing their interpretation.

## Experimental adaptive area interpolation

`method="adaptive_bary_area"` implements forward/reverse support selection and
anatomical area correction as a separate method. It requires
`experimental=TRUE`: the full-template Workbench parity gate has not passed.
There is no substitution of ordinary barycentric interpolation when it fails.

Bind areas to the exact registration geometry using `surface_vertex_areas()`.
Provide positive measured vertex areas with explicit provenance, or a
corresponding anatomical mesh with the same ordered faces. The following
ellipsoids are synthetic illustrations, not anatomical reference data.

```{r adaptive}
source_anatomy <- surface_mesh(vertices %*% diag(c(20,30,40)), faces)
target_anatomy <- surface_mesh(rotated_vertices %*% diag(c(20,30,40)), faces)
source_areas <- surface_vertex_areas(moving, anatomical=source_anatomy)
target_areas <- surface_vertex_areas(rotated, anatomical=target_anatomy)
adaptive <- surface_resampling_plan(
  rotated, moving, method="adaptive_bary_area",
  source_areas=source_areas, target_areas=target_areas, experimental=TRUE
)
result <- apply_surface_resampling(adaptive, 1+z, details=TRUE)
c(source_integral=sum(source_areas$values*(1+z)),
  effective_target_integral=sum(result$effective_target_area[result$available] *
                               result$values[result$available]))
```

Areas must be finite and strictly positive, with matching units on both sides.
Zero areas are rejected. Normalized spherical areas are never substituted
implicitly for anatomical areas.

The area identity uses an **effective target measure**, not generally the
supplied target vertex areas. With row normalization and no target exclusion,
it preserves the integral over included, represented source values. Under
omission, the effective measure is column specific and the identity includes
only finite source values; filter unavailable rows before multiplying by their
zero measure. `effective_target_area` is unavailable for `normalize="sum"` or
`"none"`. `source_roi_target_area` remains a diagnostic before data omission or
target exclusion. No conservation of numeric label IDs is claimed.

## Qualification boundaries

| Capability | Evidence and current boundary |
|---|---|
| Ordinary continuous interpolation | Analytic opposite-side, boundary and permutation tests; independent small Workbench basis fixtures; exact indexed/exhaustive comparisons. Full-template strict Workbench parity remains incomplete. |
| ROI, missing data and categorical voting | Independent policy fixtures and small Workbench mask/non-tied-label comparisons. Tie rules and missing-label keys must be mapped explicitly. |
| Reverse plans and Euclidean adjoints | Independent dense-matrix identities and rebuilt geometric reverse tests. Neither recovers discarded information. |
| Adaptive area interpolation | Analytic area/ROI identities and unequal-area Workbench basis fixtures. Experimental until full-template gates pass. |

The full ordinary-template run found rare near-edge differences above its
predeclared `5e-5` maximum-error threshold. Independent closest-point calculations
support the native result at those queries; Workbench selects an edge and omits
a small positive third weight. The failed threshold is retained. Passing a
constant field or a small fixture does not qualify every production route.

Versioned receipts and the method capability matrix are in the repository's
`output/surface-completion/` directory. Workbench is an optional comparator, not
a runtime dependency. Its [metric](https://www.humanconnectome.org/software/workbench-command/-metric-resample)
and [label](https://www.humanconnectome.org/software/workbench-command/-label-resample)
commands define the external comparison methods.
